---
title: "Project Notebook"
author: "by MoonJam"
output: rmarkdown::github_document

---

## Introduction
In the spirit of J. Allen Hynek, now that the data are processed - it's have a look at what we can learn about the reported UFO sightings in the last several decades. For the purpose of this exploration, I'm maintaining a scope of 1947 to present - using the year of the Roswell incident as our year one.

### Limitations
I should probably set this straight from the beginning: I am not purporting to answer any questions, but rather just take the available data and see what can be understood from the source as collected. I do not have any means of validating the collected data nor do I have any connection to NUFORC or any other UFO-related organization. I'm just a guy who works with data and thinks this is a cool and interesting topic.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(tidytext)
library(wesanderson)
ufo <- read_csv("ufo_data.csv")

royal1 <- wes_palette("Royal1")
royal2 <- wes_palette("Royal2")

```

```{r, echo=FALSE}
# Definitions for visualizations that will be used frequently will help keep things succinct down the road.
column_chart <- function(df, aggregate_function, index, metric, title, color=royal1[2], group=FALSE, top_n=10, by_index=TRUE) {
  if (group == TRUE) {
    df <- df %>%
      group_by_at(vars(index)) %>%
      summarize_at(vars(metric), aggregate_function)
  }
  
  if (by_index == TRUE) {
    factor_levels <- df %>%
      arrange_at(vars(metric), desc) %>%
      select(index) %>% 
      unlist(use.names = FALSE)
  
    df <- df %>%
      mutate_at(vars(index), factor, ordered = TRUE, levels = factor_levels) %>% 
      arrange_at(vars(index))
  } else {
    df <- df %>% 
      arrange_at(vars(metric), desc)
  }
  
  df %>%
    head(top_n) %>% 
    ggplot(aes_string(x = index, y = metric)) +
    geom_col(fill = color) +
    ggtitle(title)
  
}


histogram <- function(df, variable, fill, bins = 30) {
  ggplot(df, aes_string(variable)) +
    geom_histogram(fill = fill, bins = bins) +
    xlab(paste0(variable, " of UFO Sightings (in seconds)")) +
    ylab("Number of Reported UFO Sightings") +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(labels = scales::comma)
}

```

## Initial analysis
Starting with a basic glance at variable summaries, there is at least one absurd outlier for the duration of a sighting. Because I processed the data myself, I am aware of what variables are non-text and did not choose to inspect the variables' types. In the very least, we can at least verify that reports sourced from MUFON account for a negligible amount of the data.
```{r echo=FALSE}
ufo %>% select("Date/Time", Duration, "MUFON Report") %>% summary()

```

## UFO Sightings: When?
In *The UFO Experience*, J. Allen Hynek states that UFO sightings in the 1960s/1970s were on the rise. Recently, I have come across a number of references to the lack of modern sightings in various comments, threads, etc. for UFO-related media. Observing the number of sightings over time, we can see that Hynek's assertion seems reasonable considering a publication date of 1972 for *The UFO Experience*. However, it looks like there were two flaps occuring after 1994 and 2010 and subsequently a steep decline since 2014. Despite this, it seems that the last decade has had more reported UFO sightings than back in Hynek's day. This may also be due to the general availability of something like NUFORC over the last few decades relative to the 1960s/1970s - we have more documented sightings **because** there of access to a reporting center.
```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>%
  mutate(Year = year(`Date/Time`)) %>% 
  group_by(Year) %>% 
  summarize(Sightings = n()) %>%
  ggplot(aes(Year, Sightings)) +
  geom_area(size = 1, fill = royal2[1], alpha = 0.6) +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Timeline of Annual UFO Sightings")
  
```

Knowing the trend in yearly sightings has changed substantially over the last few decades, I thought it wise to inspect the distribution of sightings by year. Limiting myself to just the United States (more on this in a later section), we can see that the data fall in a non-normal distribution. The bottom 50% of observed years reach about 242 sightings with the bottom 75% hitting about 2,714 sightings. Observing and quantifying the non-normal distribution below, it is a safer bet to use the median as the measure of central tendency, which again lands us at about 242 sightings in a typical year in the U.S. alone.
```{r, echo=FALSE, fig.width=9, fig.length=9}
full_year_sightings <- ufo %>%
  filter(Country == "United States of America") %>% 
  group_by(Year = year(`Date/Time`)) %>%
  summarize(n_sightings = round(n())) %>% 
  ungroup() %>% 
  filter(Year != 2019)

full_year_sightings %>%
  ggplot(aes("x", n_sightings)) +
  geom_boxplot(fill = royal1[2]) +
  ylab("Sightings Reported by Year") +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  ggtitle("Distribution of Annual UFO Sightings")

```

All of this put together, we can see that the period of time that began in the mid-1990s really was quite exceptional by any stretch. This may be chalked up to the ease of reporting as people gained more ready access to home computers and NUFORC itself became easier to reach through the internet whenever its website went live. Although, I don't know that this really accounts for the steep decline in more recent years. Factors like public awareness of NUFORC probably roll into this as well.

If we look at what time of day these sightings are occurring, we see that about 25% would qualify as one of Hynek's daylight disks - the vast majority of sightings take place in the low visibility of night hours. This fits the generally accepted image most people have in mind when thinking of a UFO sighting.
```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>%
  mutate(Daytime = am(`Date/Time`), Nighttime = !Daytime) %>%
  summarize_at(c("Daytime", "Nighttime"), mean) %>% 
  gather("Occurrence", "Percentage", Daytime, Nighttime) %>% 
  ggplot(aes(Occurrence, Percentage, fill = Occurrence)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "none") +
  scale_fill_manual(values = royal1) +
  ggtitle("UFO Sightings by Time of Day")


```

Additionally, we can see below that separating the dates by quarters (starting with January), most sightings are occurring in the months of July, August and September. Because (as seen in the "where" section below) we know that the overwhelming majority of these recorded sightings are in North America, there isn't very much confusion going on as a result of hemispheric differences. Meaning, those months are the Summer season, and people are probably outside more often and thus more likely to see something strange in the skies.
```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>%
  mutate(Quarter = quarter(`Date/Time`), Count = 1) %>%
  group_by(Quarter) %>%
  summarize(Sightings = n()) %>%
  # mutate_at("Quarter", factor, ordered = TRUE) %>%
  ggplot(aes(Quarter, Sightings)) +
  geom_line(color = royal1[2], size = 1.5) +
  ggtitle("UFO Sightings by Quarter") +
  scale_fill_manual()

```

Generally, we see that the likelihood of seeing a UFO escalates from 6:00pm onwards at a very accelerated rate. The peak time is somewhere between 8:00pm and 10:00pm, with sightings sharply declining after 10:00pm.
```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>%
  mutate(Hour = hour(`Date/Time`)) %>%
  group_by(Hour) %>%
  summarize(Sightings = n()) %>%
  ggplot(aes(Hour, Sightings)) +
  geom_area(alpha = 0.65, fill = royal2[2], size = 1.5) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24), labels = c("12:00 AM", "2:00 AM", "4;00 AM", "6:00 AM", "8:00 AM", "10:00 AM",
                                                                                           "12:00 PM", "2:00 PM", "4:00 PM", "6:00 PM", "8:00 PM", "10:00 PM", "")) +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Hourly UFO Sightings")
```

## UFO Sightings: How?

The data includes a number of variables useful to answering questions that float around the "how" of UFO sightings. In this case, I will start with the most easily accessible which is data on how long witness report seeing their UFOs. Generally speaking, these events are described in the popular imagination as confusing, fleeting experiences and I do not anticipate lengthy time frames for most observations.

### Inspecting (non-missing) time/durations of sightings
In fact, we can see that there are a small number of reports (~ 2.7%) in the processed data that are longer than an hour. These will be put aside for later work as they are exceptional and make getting a handle of the remaining 97.3% of records difficult. A cursory glance at the isolated data suggests that missing time experiences are also being reported, but those can be explored later on as those begin to hedge on the level of the Hynek Scale's CE-III.
```{r, echo=FALSE}
over_an_hour <- ufo %>%
  mutate(over = Duration > 3600) %>%
  summarize_at("over", mean) %>%
  unlist(use.names = FALSE)
```

```{r, echo=FALSE}
over_hour <- ufo %>% filter(Duration > 3600)
ufo <- ufo %>% filter(Duration <= 3600)
```

Generating a histogram based on the 97% of processed UFO sightings, we see the duration of these events is highly right-skewed. 
```{r, echo=FALSE, fig.width=9, fig.length=9}
histogram(ufo, "Duration", royal1[1]) +
  ggtitle("Distribution of UFO Sightings by Duration")
```

Considering that any particular observation was the result of an eye witness' reporting an unexpected event, the duration variable can be thought of as discrete. Very few participants likely timed their sighting in the moment and so we can assume that these are likely estimated, general figures. While not a histogram, we see below the rate at which any particular estimated time, in seconds, was reported to NUFORC for about 90% of the processed data. From this, we can say that just over half of all sightings (54%) do not last more than 3 minutes. Additionally, only 25% ever go over 15 minutes.

```{r, echo=FALSE, fig.width=9, fig.length=9}
current_p <- 0
target_p <- 0.9

n_limit <- 15
total_n <- nrow(ufo)

while (current_p < target_p) {
  top_durations <- ufo %>% 
    group_by(Duration) %>% 
    summarize(p = round((n() / total_n), 2)) %>%
    arrange(desc(p)) %>%
    head(n_limit)
  
  current_p <- sum(top_durations$p)

  if (current_p < target_p) {
    n_limit <- n_limit + 1
  }
}

top_durations <- top_durations %>% mutate_at("p", ~ round(.x * target_p, 2))

top_durations %>%
  mutate(threeUnder = Duration <= 180, Duration = as_factor(Duration)) %>% 
  arrange(Duration) %>% 
  ggplot(aes(Duration, p, fill = threeUnder)) +
  geom_col() +
  scale_fill_manual(values=royal1) +
  theme(legend.position = "none") +
  ylab("Proportion of Sightings") +
  xlab("Duration of UFO Sightings (in seconds)") +
  ggtitle("Duration of UFO Sightings")

```

The results of this section confirm the typical narrative of a brief glimpse at a strange thing in the night sky, which even if mundane in most cases was not viewed long enough for an observer to positively identify as such.

## UFO Sightings: Where?
Considering NUFORC is based in the U.S., it may not be a surprise that we see the overwhelming majority of sightings originate from there as well. However, seeing how so few reports come in from outside the U.S., I will assume that this is because there are either other entities handling similar services abroad or that UFO-people outside of the U.S. are generally less aware of/less likely to contact NUFORC. I will revist this to see why values for countries outside of the U.S. and Canada are not currently making it through my processing step, although for now it is not a high priority given the overwhelming majority of observations are traceable to the U.S. or Canada.

```{r, echo=FALSE, fig.width=9, fig.length=9}

ufo <- ufo %>% replace_na(list("Country" = "Other Countries"))

ufo %>%
  mutate_at("Country", factor, ordered = TRUE, levels = c("United States of America", "Other Countries", "Canada")) %>% 
  ggplot(aes(Country, fill = Country)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values=royal2) +
  ylab("Number of UFO Sightings") +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  ggtitle("UFO Sightings by Country/Region")
  
```

Looking a little closer at the U.S., we can see the all-time top ten states for total UFO Sightings. Fitting with at least my own preconceptions, California looks to be the clear leader in UFO sightings. I'm a little surprised not to see more of the Southwest.
```{r, echo=FALSE, fig.width=9, fig.length=9}
state_sightings <- ufo %>%
  filter(Country == "United States of America") %>% 
  group_by("Year" = year(`Date/Time`), State) %>%
  summarize(Sightings = n()) %>%
  arrange(desc(Sightings))

column_chart(state_sightings, sum, "State", "Sightings", "Top 10 States by All-Time UFO Sightings", group=TRUE)

```
Inspecting the last full year recorded, we see that California is still the clear leader though by a less significant margin.
```{r, echo=FALSE, fig.width=9, fig.length=9}
column_chart(state_sightings %>% filter(Year == 2018), sum, "State", "Sightings", "Top 10 States in UFO Sightings During 2018", group=TRUE) +
  ggtitle("Top Ten States by UFO Sightings in 2018")
```

I need to put a bit more work into this particular theme - I feel there is a lot of insight that could be gathered by incorporating outside data sources - the presence of Airforce bases state-by-state would be something useful to use in seeking a potential correlation.

## UFO Sightings: What?

Having looked at when, where and how these sightings are occuring we're left with really looking at what exactly the sightings themselves are. The raw data contains quite a bit to work through, so I'll start with the most readily available - the UFO's shape. As processed, the data has 22 distinct possible shapes a recorded sighting might take - that's quite a few and actually represents a reduced number from raw. Alternative spellings and very similar descriptions were merged as part of my post-scrape cleaning procedures.

```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>% 
  group_by(Shape) %>% 
  summarize(Sightings = n()) %>% 
  ungroup %>% 
  column_chart(sum, "Shape", "Sightings", "UFO Shape by Number of Sightings", color = royal2[3]) +
  scale_y_continuous(labels = scales::comma) +
  coord_flip()
```

The visualization above shows a few things: first, that most UFOs can be simply described as a light in the sky or a round object. Second, is that the labels recorded by NUFORC are not hugely useful. Labels like triangular, disk and oval, which are all very distinct descriptions or shapes, come far less often than the simple "other." Given the fact that we know 75% of UFO sightings occur at night and are likely to be in sight for three minutes or less we can assume people are not getting a great look at these things. As only 25% occur during the day, it stands to reason that we have way fewer sightings where someone reports a definite shape.

```{r, echo=FALSE}
ufo <- ufo %>% mutate(Document = row_number())

doc_shape <- ufo %>% 
  select(Document, Summary) %>% 
  unnest_tokens(word, Summary) %>% 
  filter(!str_detect(word, "[\\d]") & word != "ufo") %>%
  anti_join(stop_words, by = "word") %>%
  mutate_at("word", str_replace_all, "(?<!s)s$|[:punct:]", "") %>%
  filter(word != "") %>% 
  group_by(Document, word) %>% 
  summarize(n = n())

```

Thankfully, we can actually use the summary field containing the witness' text description of the event to gather more information. Processing the text moderately, we can see what the most common words used in all recorded sightings have been - taking a bit of a leap, we can generally say that they describe "bright orange lights moving in the sky." Interestingly, in a lot of the media surrounding UFOs, they color orange is held as the standard and it certainly seems to pan out here. This also confirms a bit of what we've seen in the labels - these aren't so much descriptions of objects as much as just the lights themselves.
```{r, echo=FALSE, fig.width=9, fig.length=9}
overall_words <- doc_shape %>% 
  group_by(word) %>% 
  summarize_at("n", sum) %>%
  arrange(desc(n))

factor_levels <- overall_words$word

overall_words <- overall_words %>% mutate_at("word", factor, ordered = TRUE, levels = factor_levels)

overall_words %>%
  column_chart(sum, "word", "n", "Most Common Words Describing a UFO", royal2[2], top_n = 20) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  xlab("Word") +
  ylab("Occurrence")
```

