---
title: "Project Notebook"
author: by MoonJam
output: html_notebook
---

## Introduction
In the spirit of J. Allen Hynek, now that the data are processed - it's have a look at what we can learn about the reported UFO sightings in the last several decades. For the purpose of this exploration, I'm maintaining a scope of 1947 to present - using the year of the Roswell incident as our year one.

### Limitations
I should probably set this straight from the beginning: I am not purporting to answer any questions, but rather just take the available data and see what can be understood from the source as collected. I do not have any means of validating the collected data nor do I have any connection to NUFORC or any other UFO-related organization. I'm just a guy who works with data and thinks this is a cool and interesting topic.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(wesanderson)
ufo <- read_csv("ufo_data.csv")

royal1 <- wes_palette("Royal1")
royal2 <- wes_palette("Royal2")

```

```{r, echo=FALSE}
histogram <- function(df, variable, fill, bins = 30) {
  ggplot(df, aes_string(variable)) +
    geom_histogram(fill = fill, bins = bins) +
    xlab(paste0(variable, " of UFO Sightings (in seconds)")) +
    ylab("Number of Reported UFO Sightings") +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(labels = scales::comma)
}

```

## Initial analysis
Starting with a basic glance at variable summaries, there is at least one absurd outlier for the duration of a sighting. Because I processed the data myself, I am aware of what variables are non-text and did not choose to inspect the variables' types. In the very least, we can at least verify that reports sourced from MUFON account for a negligible amount of the data.
```{r echo=FALSE}
ufo %>% select("Date/Time", Duration, "MUFON Report") %>% summary()
```

## UFO Sightings: When?
In *The UFO Experience*, J. Allen Hynek states that UFO sightings in the 1960s/1970s were on the rise. Recently, I have come across a number of references to the lack of modern sightings in various comments, threads, etc. for UFO-related media. Observing the number of sightings over time, we can see that Hynek's assertion seems reasonable considering a publication date of 1972 for *The UFO Experience*. However, it looks like there were two flaps occuring after 1994 and 2010 and subsequently a steep decline since 2014. Despite this, it seems that the last decade has had more reported UFO sightings than back in Hynek's day.
```{r, echo=FALSE, fig.width=9, fig.length=9}
ufo %>%
  mutate(Year = year(`Date/Time`)) %>% 
  group_by(Year) %>% 
  summarize(Sightings = n()) %>%
  ggplot(aes(Year, Sightings)) +
  geom_area(size = 1, fill = royal2[1], alpha = 0.6) +
  scale_y_continuous(labels = scales::comma)
  
```

Knowing the trend in yearly sightings has changed substantially over the last few decades, I thought it wise to inspect the distribution of sightings by year. Limiting myself to just the United States (more on this in a later section), we can see that the data fall in a non-normal distribution. The bottom 50% of observed years reach about 242 sightings with the bottom 75% hitting about 2,714 sightings. Observing and quantifying the non-normal distribution below, it is a safer bet to use the median as the measure of central tendency, which again lands us at about 242 sightings in a typical year in the U.S. 
```{r, echo=FALSE, fig.width=9, fig.length=9}
full_year_sightings <- ufo %>%
  filter(Country == "United States of America") %>% 
  group_by(Year = year(`Date/Time`)) %>%
  summarize(n_sightings = round(n())) %>% 
  ungroup() %>% 
  filter(Year != 2019)

full_year_sightings %>%
  ggplot(aes("x", n_sightings)) +
  geom_boxplot(fill = royal1[2]) +
  ylab("Sightings Reported by Year") +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())

```

All of this put together, we can see that the period of time that began in the mid-1990s really was quite exceptional by any stretch. This may be chalked up to the ease of reporting as people gained more ready access to home computers and NUFORC itself became easier to reach through the internet whenever its website went live. Although, I don't know that this really accounts for the steep decline in more recent years. Even isolating to just the available months of the current year, it looks like every year since at least 2015 has shown a steady decline of reported sightings.

## UFO Sightings: How?

The data includes a number of variables useful to answering questions that float around the "how" of UFO sightings. In this case, I will start with the most easily accessible which is data on how long witness report seeing their UFOs. Generally speaking, these events are described in the popular imagination as confusing, fleeting experiences and I do not anticipate lengthy time frames for most observations.

### Inspecting (non-missing) time/durations of sightings
In fact, we can see that there are a small number of reports (~ 2.7%) in the processed data that are longer than an hour. These will be put aside for later work as they are exceptional and make getting a handle of the remaining 97.3% of records difficult. A cursory glance at the isolated data suggests that missing time experiences are also being reported, but those can be explored later on as those begin to hedge on the level of the Hynek Scale's CE-III.
```{r, echo=FALSE}
over_an_hour <- ufo %>%
  mutate(over = Duration > 3600) %>%
  summarize_at("over", mean) %>%
  unlist(use.names = FALSE)

print(paste0("Sightings reported at over an hour: ", round(over_an_hour * 100, 2), "%"))
```
```{r, echo=FALSE}
over_hour <- ufo %>% filter(Duration > 3600)
ufo <- ufo %>% filter(Duration <= 3600)
```

Generating a histogram based on the 97% of processed UFO sightings, we see the duration of these events is highly right-skewed. 
```{r, echo=FALSE, fig.width=9, fig.length=9}
histogram(ufo, "Duration", royal1[1])
```

Considering that any particular observation was the result of an eye witness' reporting an unexpected event, the duration variable can be thought of as discrete. Very few participants likely timed their sighting in the moment and so we can assume that these are likely estimated, general figures. While not a histogram, we see below the rate at which any particular estimated time, in seconds, was reported to NUFORC for about 90% of the processed data. From this, we can say that just over half of all sightings (54%) do not last more than 3 minutes. Additionally, only 25% ever go over 15 minutes.

```{r, echo=FALSE, fig.width=9, fig.length=9}
current_p <- 0
target_p <- 0.9

n_limit <- 15
total_n <- nrow(ufo)

while (current_p < target_p) {
  top_durations <- ufo %>% 
    group_by(Duration) %>% 
    summarize(p = round((n() / total_n), 2)) %>%
    arrange(desc(p)) %>%
    head(n_limit)
  
  current_p <- sum(top_durations$p)

  if (current_p < target_p) {
    n_limit <- n_limit + 1
  }
}

top_durations <- top_durations %>% mutate_at("p", ~ round(.x * target_p, 2))

top_durations %>%
  mutate(threeUnder = Duration <= 180, Duration = as_factor(Duration)) %>% 
  arrange(Duration) %>% 
  ggplot(aes(Duration, p, fill = threeUnder)) +
  geom_col() +
  scale_fill_manual(values=royal1) +
  theme(legend.position = "none") +
  ylab("Proportion of Sightings") +
  xlab("Duration of UFO Sightings (in seconds)")

```

For now, I'm pretty satisfied in this section. The results confirm the typical narrative of a brief glimpse at a strange thing in the sky, which even if mundane in most cases was not viewed long enough for an observer to positively identify as such. I think this variable coupled with the summary variable that contains the observer's description of the event will prove pretty interesting whenever I get to it. 

## UFO Sightings: Where?
Considering NUFORC is based in the U.S., it may not be a surprise that we see the overwhelming majority of sightings originate from there as well. However, seeing how so few reports come in from outside the U.S., I will assume that this is because there are either other entities handling similar services abroad or that UFO-people outside of the U.S. are generally less aware of/less likely to contact NUFORC. I will revist this to see why values for countries outside of the U.S. and Canada are not currently making it through my processing step, although for now it is not a high priority given the overwhelming majority of observations are traceable to the U.S. or Canada.

```{r, echo=FALSE, fig.width=9, fig.length=9}

ufo <- ufo %>% replace_na(list("Country" = "Other Countries"))

ufo %>%
  mutate_at("Country", factor, ordered = TRUE, levels = c("United States of America", "Other Countries", "Canada")) %>% 
  ggplot(aes(Country, fill = Country)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values=royal2) +
  ylab("Number of UFO Sightings") +
  theme(legend.position = "none", axis.title.x = element_blank())
  
```

